---
title: "EDA"
author: "Code Notebook"
date: "2025-04-21"
output: html_document
---


```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(ggplot2)
library(scales)
library(rlang)
```



```{r}
setwd("C:/Users/dylan/Downloads/Data 205/Water Use")
water_use <- read_csv("Water_Use_10States.csv") %>% clean_names()
tewu <- read_csv("TEWU_10States.csv") %>% clean_names()
population <- read_csv("Population_data.csv") %>% clean_names()
irrigation <- read_excel("Irrigation.xlsx") %>% clean_names()
water_quality <- read_csv("BKB_WaterQualityData_2020084.csv") %>% clean_names()
```



```{r}
state_map <- tibble(
  state = c("CA", "WA", "UT", "SD", "TX", "IL", "FL", "MD", "ME", "AZ"),
  year_col = c("year_2", "year_5", "year_8", "year_11", "year_14",
               "year_17", "year_20", "year_23", "year_26", "year_29"),
  mgd_col  = c("pscutot_mgd_ca", "pscutot_mgd_wa", "pscutot_mgd_ut", "pscutot_mgd_sd",
               "pscutot_mgd_tx", "pscutot_mgd_il", "pscutot_mgd_fl", "pscutot_mgd_md",
               "pscutot_mgd_me", "pscutot_mgd")
)

# Build tidy long-format dataframe
water_use_long <- purrr::map_dfr(1:nrow(state_map), function(i) {
  state <- state_map$state[i]
  y_col <- state_map$year_col[i]
  m_col <- state_map$mgd_col[i]
  
  tibble(
    state = state,
    year = as.numeric(water_use[[y_col]]),
    mgd = as.numeric(water_use[[m_col]])
  ) %>% drop_na()
})
```
This reshapes a wide-format dataset (in this case, water_use) into a tidy long-format dataframe called water_use_long. It starts by creating a mapping table (state_map) that links each U.S. state's abbreviation to its corresponding year and water use (MGD) column names. Then, it uses purrr::map_dfr() to loop over each row in that mapping table. For each state, the function extracts the appropriate year and mgd (millions of gallons per day) values from the original dataset, combines them into a new tibble with the state name, and removes rows with missing data using drop_na(). The final output is a clean long-format dataframe where each row represents a specific state-year-MGD combination, which is ideal for plotting and further analysis.


```{r}
tewu_long <- tewu %>%
  pivot_longer(cols = starts_with("huc12_id"), names_to = "huc12_id_var", values_to = "huc12_id") %>%
  bind_cols(
    tewu %>%
      pivot_longer(cols = starts_with("year"), names_to = "year_var", values_to = "year") %>%
      select(year),
    tewu %>%
      pivot_longer(cols = starts_with("tecuftot_mgd"), names_to = "tecuftot_mgd_var", values_to = "tecuftot_mgd") %>%
      select(tecuftot_mgd)
  ) %>%
  drop_na(year)
```
This R code transforms the `tewu` dataset (thermoelectric water use data) from wide to long format using a combination of pivot_longer() and bind_cols(). The goal is to align each thermoelectric water use value (tecuftot_mgd) with its corresponding year and HUC12 geographic identifier. First, it reshapes all huc12_id_* columns into a single huc12_id column while tracking their original variable names in huc12_id_var. Next, it binds this to a reshaped version of the year columns (e.g., year`, year.1, etc.) and another reshaped version of the tecuftot_mgd_* columns. Each of these transformations extracts the values into a shared long structure so that each row represents a unique combination of location, year, and water use amount. Finally, it filters out any rows where year is missing using drop_na(year). This format is better suited for analysis and visualization.


```{r}
population <- population %>%
  mutate(total_pop = rowSums(select(., starts_with("pop_")), na.rm = TRUE))

population_clean <- population %>%
  mutate(
    sex = recode(sex_1, `0` = "Both", `1` = "Male", `2` = "Female"),
    origin = recode(origin_2, `0` = "Total", `1` = "Not Hispanic", `2` = "Hispanic"),
    race = recode(race_3, `0` = "All", `1` = "White", `2` = "Black", `3` = "AIAN",
                  `4` = "Asian", `5` = "NHPI", `6` = "2+ Races", `7` = "White Combo",
                  `8` = "Black Combo", `9` = "AIAN Combo", `10` = "Asian Combo", `11` = "NHPI Combo")
  )
```
This R code prepares the population dataset for analysis by aggregating total population and converting coded variables into readable labels. First, it creates a new column, total_pop, by summing across all columns that begin with `"pop_"`, which represent different age groups or population categories; na.rm = TRUE ensures missing values are ignored during summation. Next, the population_clean object is created by recoding numeric codes from the original columns into descriptive categories: sex_1 is transformed into sex with values "Male", "Female", and "Both"; origin_2 into origin as "Hispanic", "Not Hispanic", or "Total"; and race_3 into race, which includes categories such as "White", "Black", "Asian", and several multi-race combinations. These transformations improve readability and make the dataset more suitable for visualizations and demographic analysis.


```{r}
water_quality_clean <- water_quality %>%
  filter(!is.na(site_id), !is.na(year))
```
This R code filters the water_quality dataset to create a cleaner version named water_quality_clean by removing rows with missing values in key columns. Specifically, it uses the filter() function to retain only those observations where both site_id and year are not NA. These two fields are typically essential for temporal and spatial analysis, as site_id identifies the monitoring location and year indicates when the measurements were taken. By ensuring both are present, the code improves the reliability and integrity of any analysis or visualization that depends on those variables.



```{r}
state_map <- tibble(
  state = c("CA", "WA", "UT", "SD", "TX", "IL", "FL", "MD", "ME", "AZ"),
  year_col = c("year_2", "year_5", "year_8", "year_11", "year_14",
               "year_17", "year_20", "year_23", "year_26", "year_29"),
  mgd_col  = c("pscutot_mgd_ca", "pscutot_mgd_wa", "pscutot_mgd_ut", "pscutot_mgd_sd",
               "pscutot_mgd_tx", "pscutot_mgd_il", "pscutot_mgd_fl", "pscutot_mgd_md",
               "pscutot_mgd_me", "pscutot_mgd")
)

# Build long-format tidy irrigation dataset
irrigation_long <- purrr::map_dfr(1:nrow(state_map), function(i) {
  s <- state_map$state[i]
  y_col <- state_map$year_col[i]
  m_col <- state_map$mgd_col[i]
  
  tibble(
    state = s,
    year = as.numeric(irrigation[[y_col]]),
    mgd = as.numeric(irrigation[[m_col]])
  ) %>% drop_na()
})
```
This R code reshapes a wide-format irrigation dataset into a long-format structure suitable for analysis and visualization. It begins by defining a state_map tibble that associates each state abbreviation with its corresponding year and irrigation water use (mgd) column names in the dataset. Using purrr::map_dfr(), the code loops over each row of state_map, dynamically pulling the appropriate columns from the original irrigation dataset for each state. For every iteration, it creates a tibble containing the state name, the year (converted to numeric), and the water use in million gallons per day (mgd). The drop_na() function is applied to remove rows with missing values in either the year or mgd columns. The final output, irrigation_long, is a tidy dataframe with one observation per state-year-MGD combination, making it ideal for time-series plots or comparative analysis across states.



```{r}
summary(water_use_long$pscutot_mgd)
summary(tewu_long$tecuftot_mgd)
summary(water_quality_clean$salinity_ppt)
summary(population_clean$total_pop)
```


## Water Use
```{r}
ggplot(water_use_long, aes(x = year, y = mgd)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Public Supply Water Use by State (MGD)",
    x = "Year",
    y = "Million Gallons per Day"
  ) +
  theme_minimal()

ggplot(water_use_long, aes(x = mgd)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Water Use (MGD)", x = "MGD", y = "Count") +
  theme_minimal()
```
Graph 1:
This histogram visualizes the distribution of water use across all observations in million gallons per day (MGD). The chart is highly right-skewed, with the vast majority of values concentrated near zero. Most observations fall below 1 MGD, as indicated by the tall bar on the left, and frequency rapidly declines as usage increases. This suggests that low water usage is far more common across the dataset, while high-consumption cases are rare and appear as long-tail outliers. Such a distribution may be typical of residential or localized public supply systems, where only a few locations demand large volumes of water.

Graph 2:
The second graph is a duplicate of the first and presents the same histogram showing the distribution of water use in MGD. The pattern again emphasizes extreme skewness, where water use is minimal in most cases and only a small fraction of observations exceed a few MGD. This reinforces the insight that most water users or sites consume relatively little water, and helps identify the need for possible log transformation if statistical modeling is planned. Repeated visualization also helps validate data consistency and the presence of extreme values that may need further investigation.

## Thermo Electric Water Use
```{r}
ggplot(tewu_long, aes(x = year, y = tecuftot_mgd)) +
  geom_line(stat = "summary", fun = mean, color = "orange") +
  labs(title = "Average Thermoelectric Water Use Over Time", x = "Year", y = "MGD") +
  theme_minimal()

ggplot(tewu_long, aes(x = factor(year), y = tecuftot_mgd)) +
  geom_boxplot(fill = "coral", outlier.alpha = 0.2) +
  labs(title = "Distribution of Thermoelectric Use by Year", x = "Year", y = "MGD") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```
Graph 1:
This line graph illustrates the trend in average thermoelectric water use (measured in million gallons per day, or MGD) from 2008 to 2020. Each point represents the mean water usage for a given year, and the orange line connects these yearly averages to show the overall trajectory. The graph reveals a notable decline in average usage between 2008 and 2011, hitting a low around 2011. After that, usage begins to gradually increase, with a more pronounced upward trend from 2017 to 2020, peaking in 2020. This pattern may reflect changes in thermoelectric technology, regulation, climate conditions, or energy demand over time. The relatively low y-axis values indicate that while usage is increasing, average MGD remains under 0.05 across all years, suggesting generally low per-site consumption in this category.

Graph 2:
This scatter plot displays the distribution of thermoelectric water use in million gallons per day (MGD) from 2008 to 2020. Each point represents a unique observation in a specific year, allowing for a visual assessment of variability and trends over time. The data is heavily concentrated near the lower end of the MGD scale, indicating that most thermoelectric sites use relatively small amounts of water. However, occasional outliers reach values above 5 or even 8 MGD in certain years. There does not appear to be a strong upward or downward trend across the timeframe, but the consistent pattern of low values with sparse high outliers suggests the need for further investigation into what types of facilities or conditions produce higher usage.


## Population
```{r}
plot_data <- population_clean %>%
  filter(sex == "Both", origin == "Total", race == "All")

# Plot total population over time
ggplot(plot_data, aes(x = year_4, y = total_pop)) +
  geom_line(color = "darkgreen") +
  labs(
    title = "Projected Total US Population",
    x = "Year",
    y = "Total Population"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```
This line graph presents the projected total population of the United States from 2020 through 2100. The y-axis represents total population in raw numbers (with comma separators for readability), and the x-axis spans the projection years. The trend shows a steady population increase from 2020 to approximately 2080, where it peaks just under 740 million people. After reaching this apex, the population begins to gradually decline, suggesting that long-term demographic shifts such as declining birth rates, aging populations, or immigration policy changes may play a significant role in the latter half of the century. This type of visualization is critical for long-term planning in areas like infrastructure, environmental policy, healthcare, and resource allocation. The smoothness of the curve also suggests the use of modeled projections rather than year-by-year survey data.


## Irrigation Data
```{r}
ggplot(irrigation_long, aes(x = year, y = mgd)) +
  geom_line(color = "seagreen") +
  facet_wrap(~ state, scales = "free_y") +
  labs(
    title = "Irrigation Water Use by State (MGD)",
    x = "Year",
    y = "Million Gallons per Day"
  ) +
  theme_minimal()
```
This faceted line graph displays irrigation water use in million gallons per day (MGD) for ten U.S. states from approximately 2010 to 2020. Each subplot represents one state, with water use shown over time. States like Arizona (AZ) and California (CA) show consistently high irrigation use, peaking around 20 and 15 MGD respectively, indicating their large agricultural sectors and reliance on irrigation. Other states such as Maryland (MD) and Utah (UT) display moderate levels, while states like Maine (ME) and South Dakota (SD) show minimal water use, barely reaching above 0.01 MGD. The downward spikes seen in many states suggest either missing data or seasonal irrigation patterns that weren’t smoothed in the visualization. Overall, the graph clearly illustrates the diverse scale of irrigation needs across different regions, with western and southwestern states consuming the most water for agriculture.


## Water Quality
```{r}
ggplot(water_quality_clean, aes(x = year, y = water_temp_c)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Water Temperature Trend", x = "Year", y = "Temp (°C)") +
  theme_minimal()


ggplot(water_quality_clean, aes(x = `dissolved_oxygen_mg_l`)) +
  geom_histogram(bins = 25, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Dissolved Oxygen", x = "mg/L", y = "Count") +
  theme_minimal()
```
Graph 1:
This scatterplot with a fitted line illustrates the trend in recorded water temperatures (in °C) over a wide time range, from the late 19th century through 2020. Each black dot represents an individual observation, while the red line shows the smoothed trend (likely from a LOESS or GAM fit), with a shaded confidence interval. Early observations are sparse, but from around the year 2000 onward, data becomes much denser. The fitted curve indicates a general increase in water temperature in recent decades, which could reflect the effects of climate change, urban heat island effects, or altered hydrological patterns. A few extremely high outliers appear post-2000 and should be evaluated for measurement anomalies. This graph provides important insight into long-term thermal changes in aquatic environments.

Graph 2:
This histogram displays the distribution of dissolved oxygen concentrations across all water samples, measured in milligrams per liter (mg/L). The data shows a roughly normal (bell-shaped) distribution, with most values falling between 4 and 9 mg/L and a peak around 6–7 mg/L. Dissolved oxygen is critical for aquatic life, and this distribution suggests generally healthy oxygen levels in the sampled water bodies. However, the tail on the lower end (near 0 mg/L) may indicate occasional hypoxic conditions, which could be harmful to aquatic ecosystems. The presence of high DO values (>10 mg/L) on the right tail is rare but possible under cold, well-aerated conditions. This visualization offers a useful overview of water quality health based on oxygen availability.

## Water Sustainability for just drinking water
```{r}
# Total water use (annual)
water_annual <- water_use_long %>%
  group_by(year) %>%
  summarise(total_mgd = sum(mgd, na.rm = TRUE)) %>%
  mutate(total_gal_year = total_mgd * 1e6 * 365)

# Fit linear model
water_model <- lm(total_gal_year ~ year, data = water_annual)

# Create future years to 2100
future_years <- tibble(year = 2021:2100)

# Predict future water use
water_future <- future_years %>%
  mutate(total_gal_year = predict(water_model, newdata = future_years))

# Combine with historical data
water_projected <- bind_rows(water_annual, water_future)

# Population data 
pop_projected <- population_clean %>%
  filter(sex == "Both", origin == "Total", race == "All") %>%
  group_by(year = year_4) %>%
  summarise(total_pop = sum(total_pop, na.rm = TRUE)) %>%
  filter(year >= 2000, year <= 2100)

# Merge Water use and population
water_vs_population <- left_join(pop_projected, water_projected, by = "year") %>%
  mutate(
    per_capita_gallons = total_gal_year / total_pop,
    status = ifelse(total_gal_year > max(water_annual$total_gal_year, na.rm = TRUE), "unsustainable", "sustainable")
  )

#Find Unsustainable year
first_unsustainable_year <- water_vs_population %>%
  filter(status == "unsustainable") %>%
  slice(1) %>%
  pull(year)
```
This R code projects future public water use and compares it to population growth to determine when water demand becomes unsustainable. It begins by aggregating total annual public water use from the water_use_long dataset, converting daily use in million gallons per day (MGD) into total gallons per year. A linear regression model (`water_model`) is then fitted to this historical water use data to capture the overall trend.

Using this model, the code generates future predictions for water demand from 2021 through 2100, storing the results in water_future. These forecasts are combined with historical values to produce a full timeline (water_projected) of both observed and projected annual water use.

The next section prepares population data from population_clean by filtering for total population across all races and both sexes, then grouping and summarizing by year. This population data is merged with the projected water use to form water_vs_population, which includes a calculation of per capita water use and an evaluation of whether the total projected water use exceeds the historical maximum (used here as a proxy for the sustainability limit    ).

Finally, the code identifies the first year where water demand becomes unsustainable by filtering for years where demand surpasses that threshold and extracting the earliest such year. This approach highlights how long the current trend in water usage can continue before exceeding sustainable supply levels.



```{r}
ggplot(water_vs_population, aes(x = year, y = total_gal_year)) +
  geom_line(color = "blue") +
  geom_hline(yintercept = max(water_annual$total_gal_year), color = "red", linetype = "dashed") +
  labs(
    title = "Projected Water Demand vs. Sustainability Threshold",
    x = "Year",
    y = "Total Gallons Per Year"
  ) +
  theme_minimal()

#Find Unsustainable year
first_unsustainable_year <- water_vs_population %>%
  filter(status == "unsustainable") %>%
  slice(1) %>%
  pull(year)
print(paste("Water becomes unsustainable in:", first_unsustainable_year))
```
This line graph visualizes projected U.S. water demand (in gallons per year) from 2020 through 2100, compared against a sustainability threshold. The blue line represents projected annual water use based on a linear model fitted to historical water use data. The red dashed line marks the current sustainability limit — in this case, defined as the historical maximum observed water use.

The projection indicates a steady and significant increase in water demand, growing from about 1.6 × 10¹¹ gallons/year in 2020 to over 3.0 × 10¹¹ gallons/year by 2100. This upward trend reflects the influence of population growth and constant or increasing per capita usage.

Critically, the graph shows that the blue line crosses the sustainability threshold early in the projection period, indicating that water use becomes unsustainable within the next few decades — unless conservation, efficiency, or supply expansion measures are implemented. This visualization highlights an urgent need for strategic water resource planning in the face of rising population pressure and potential climate change impacts.


## Water Sustainability for TEWU, Irrigation, and Public Supply
```{r}
# Preparing all Three Data sets
public_use <- water_use_long %>%
  group_by(year) %>%
  summarise(public_gal = sum(mgd, na.rm = TRUE) * 1e6 * 365)

tewu_use <- tewu_long %>%
  group_by(year) %>%
  summarise(tewu_gal = sum(tecuftot_mgd, na.rm = TRUE) * 1e6 * 365)

irrigation_use <- irrigation_long %>%
  group_by(year) %>%
  summarise(irrigation_gal = sum(mgd, na.rm = TRUE) * 1e6 * 365)

# Combine All Uses
water_total <- full_join(public_use, tewu_use, by = "year") %>%
  full_join(irrigation_use, by = "year") %>%
  arrange(year) %>%
  mutate(across(ends_with("_gal"), ~replace_na(., 0))) %>%
  mutate(total_gal_year = public_gal + tewu_gal + irrigation_gal)

# Linear model on total demand
water_model <- lm(total_gal_year ~ year, data = water_total)

# Future projection
future_years <- tibble(year = 2021:2100)
water_future <- future_years %>%
  mutate(total_gal_year = predict(water_model, newdata = future_years))

# Combine historical and projected
water_projected_all <- bind_rows(water_total %>% select(year, total_gal_year), water_future)

# Population filtered to match projection range
pop_projected <- population_clean %>%
  filter(sex == "Both", origin == "Total", race == "All") %>%
  group_by(year = year_4) %>%
  summarise(total_pop = sum(total_pop, na.rm = TRUE)) %>%
  filter(year >= 2000, year <= 2100)

# Merge and assess
comparison <- left_join(pop_projected, water_projected_all, by = "year") %>%
  mutate(
    per_capita_gal = total_gal_year / total_pop,
    sustainable_limit = max(water_total$total_gal_year, na.rm = TRUE),
    status = ifelse(total_gal_year > sustainable_limit, "unsustainable", "sustainable")
  )
```
This block of R code performs a comprehensive analysis to evaluate future water sustainability in the United States by integrating three key sources of water use — public supply, thermoelectric (TEWU), and irrigation — and comparing their combined projected demand against long-term population growth.

First, the code aggregates total annual water use for each category by converting daily usage (in million gallons per day, MGD) into yearly totals (in gallons). It then merges the three data sets into a single water_total table, replacing any missing values with zeros to ensure consistency. The total annual demand is calculated by summing water use across all three sectors for each year. A linear model is fitted to this total water demand, which is then used to project water usage through the year 2100. These projected values are combined with historical values in the water_projected_all data set.

Next, the code filters and summarizes population projections to match the years of the water demand data. It merges population and water use projections into a unified table, comparison, and calculates per capita water use for each year. A sustainability threshold is defined using the maximum historical total water use, and each year is labeled as either "sustainable" or "unsustainable" based on whether projected demand exceeds this threshold. This allows for clear identification of when water use becomes unsustainable under current growth trends.


```{r}
# Visualization
ggplot(comparison, aes(x = year, y = total_gal_year)) +
  geom_line(color = "darkorange") +
  geom_hline(aes(yintercept = sustainable_limit), linetype = "dashed", color = "red") +
  labs(
    title = "Combined Water Demand vs. Sustainability Threshold",
    x = "Year",
    y = "Total Gallons Per Year"
  ) +
  theme_minimal()

# First unsustainable year
first_unsustainable_year <- comparison %>%
  filter(status == "unsustainable") %>%
  slice(1) %>%
  pull(year)

print(paste("Water becomes unsustainable in:", first_unsustainable_year))
```
This graph projects total U.S. water demand across public supply, thermoelectric, and irrigation sources from 2020 through 2100, measured in gallons per year. The orange line represents the combined annual water use forecast, while the red dashed line marks the sustainability threshold — defined here as the historical maximum total water use observed from the combined sources.

The projection shows a steadily rising demand, increasing from just over 3.0 × 10¹¹ gallons per year in 2020 to over 1.2 × 10¹² by 2100. This dramatic climb reflects compounding growth across population and sectoral water needs. Notably, the orange line crosses the sustainability threshold early in the timeline, signaling that under current usage patterns, the U.S. will exceed sustainable water capacity within the next two decades.

This visualization underscores the urgent need for water conservation, efficiency strategies, technological innovation, and potentially revised water allocation policies to avoid long-term shortages. Let me know if you'd like to:


```{r}
public_avg <- water_use_long %>%
  summarise(avg_public_mgd = mean(mgd, na.rm = TRUE))

tewu_avg <- tewu_long %>%
  summarise(avg_tewu_mgd = mean(tecuftot_mgd, na.rm = TRUE))

irrigation_avg <- irrigation_long %>%
  summarise(avg_irrigation_mgd = mean(mgd, na.rm = TRUE))

# Combine into one summary table
daily_use_summary <- bind_cols(public_avg, tewu_avg, irrigation_avg)

print(daily_use_summary)
```
This table summarizes the average daily water use across three major sectors in the dataset, measured in million gallons per day (MGD). Both the public supply and irrigation sectors show an average use of approximately 0.51 MGD, indicating that residential and agricultural demands are the dominant contributors to overall water consumption. In contrast, thermoelectric water use is significantly lower, averaging around 0.036 MGD. While thermoelectric facilities can have high localized water demands, their overall average is much lower than the other sectors in this dataset. These figures reflect per-unit averages (such as per site or HUC region) and highlight the substantial water burden from public and agricultural use, reinforcing the importance of targeting these sectors in water conservation strategies.



```{r}
water_quality_clean <- water_quality_clean %>%
  mutate(
    pH_status = case_when(
      p_h_standard_units >= 6.5 & p_h_standard_units <= 8.5 ~ "Safe",
      TRUE ~ "Unsafe"
    ),
    salinity_status = case_when(
      salinity_ppt > 0.5 ~ "High",
      TRUE ~ "Normal"
    )
  )

ggplot(water_quality_clean, aes(x = pH_status, fill = pH_status)) +
  geom_bar() +
  labs(
    title = "pH Classification of Water Samples",
    x = "pH Status",
    y = "Number of Samples"
  ) +
  scale_fill_manual(values = c("Safe" = "seagreen", "Unsafe" = "firebrick")) +
  theme_minimal()

ggplot(water_quality_clean, aes(x = salinity_status, y = salinity_ppt, fill = salinity_status)) +
  geom_boxplot() +
  labs(
    title = "Salinity Levels in Water Samples",
    x = "Salinity Classification",
    y = "Salinity (ppt)"
  ) +
  scale_fill_manual(values = c("Normal" = "skyblue", "High" = "tomato")) +
  theme_minimal()
```
This boxplot highlights concerns about low drinkability in several water samples based on salinity levels. According to EPA and WHO guidelines, drinking water should typically have salinity levels below 0.5 parts per thousand (ppt). In the plot, samples categorized as "Normal" fall well within this safe range, with salinity values near zero. However, the "High" category shows a substantial increase in salinity, with a median around 2.5 ppt and outliers reaching over 7 ppt — far above drinkable thresholds. These elevated levels may indicate saltwater intrusion in coastal aquifers, industrial runoff, or poor water treatment and infrastructure degradation. The presence of such high salinity in a significant portion of samples suggests that many water sources in this dataset are not suitable for drinking without treatment, emphasizing the need for targeted interventions in high-risk regions.


## Water Salinity V Water Use
```{r}
salinity_summary <- water_quality_clean %>%
  group_by(year) %>%
  summarise(avg_salinity = mean(salinity_ppt, na.rm = TRUE))

water_use_summary <- water_use_long %>%
  group_by(year) %>%
  summarise(total_mgd = sum(mgd, na.rm = TRUE))

use_vs_quality <- left_join(water_use_summary, salinity_summary, by = "year")
```
salinity_summary: Calculates the average salinity (salinity_ppt) for each year from the cleaned water quality dataset (water_quality_clean). This gives an annual overview of water quality.

water_use_summary: Sums the total water use (mgd) by year from the long-format water use dataset (water_use_long). This represents the total volume of water used per year in million gallons per day.

use_vs_quality: Merges the two summaries into one dataset using a left join on the year column. The resulting dataset allows for side-by-side comparison of annual water usage and average salinity, enabling further correlation analysis or visualization.


```{r}
ggplot(use_vs_quality, aes(x = year)) +
  geom_line(aes(y = total_mgd, color = "Total Water Use (MGD)")) +
  geom_line(aes(y = avg_salinity * 1000, color = "Average Salinity (ppt × 1000)")) +
  labs(
    title = "Water Use vs. Salinity Levels Over Time",
    x = "Year",
    y = "Scaled Values",
    color = "Metric"
  ) +
  theme_minimal()

ggplot(use_vs_quality, aes(x = total_mgd, y = avg_salinity)) +
  geom_point(color = "darkred") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Correlation Between Water Use and Salinity",
    x = "Total Water Use (MGD)",
    y = "Average Salinity (ppt)"
  ) +
  theme_minimal()
```
Graph 1:
This line graph compares total water use in million gallons per day (MGD) with average salinity levels over time. Water use appears relatively stable from 2010 to 2020, with only minor fluctuations. In contrast, average salinity levels show dramatic shifts — dropping to near-zero for multiple years and then spiking sharply before declining again. The salinity line was scaled (×1000) for comparison, but the extreme variation suggests either localized environmental events or inconsistencies in the salinity data (such as missing or imputed values). Overall, the graph does not show a strong time-based relationship between general water use and salinity levels, implying that factors other than overall consumption — such as land use, contamination sources, or treatment system failures — are likely responsible for changes in water quality.

Graph 2:
This scatterplot shows the direct relationship between total water use (MGD) and average salinity (ppt). Each point represents a year. The regression line is nearly flat and slightly negative, and the shaded confidence band shows high uncertainty. This indicates no meaningful correlation between the amount of water used and salinity levels across years. While a few outliers exist — where salinity levels are unusually high despite moderate usage — these appear to be exceptions rather than evidence of a broader trend. These findings support the idea that salinity is driven more by local environmental or geological conditions than by overall water demand.

 
## Water Use V Water Quality
```{r}
public_use <- water_use_long %>%
  group_by(year) %>%
  summarise(public_gal = sum(mgd, na.rm = TRUE) * 1e6 * 365)

tewu_use <- tewu_long %>%
  group_by(year) %>%
  summarise(tewu_gal = sum(tecuftot_mgd, na.rm = TRUE) * 1e6 * 365)

irrigation_use <- irrigation_long %>%
  group_by(year) %>%
  summarise(irrigation_gal = sum(mgd, na.rm = TRUE) * 1e6 * 365)

# Combine Sectoral Data
water_combined <- reduce(
  list(public_use, tewu_use, irrigation_use),
  full_join, by = "year"
) %>%
  arrange(year) %>%
  mutate(across(ends_with("_gal"), ~replace_na(., 0))) %>%
  mutate(total_gal = public_gal + tewu_gal + irrigation_gal)

# Stacked area plot
water_long <- water_combined %>%
  select(year, public_gal, tewu_gal, irrigation_gal) %>%
  pivot_longer(cols = -year, names_to = "sector", values_to = "gallons")


sustainable_limit <- max(water_combined$total_gal, na.rm = TRUE)

ggplot(water_long, aes(x = year, y = gallons, fill = sector)) +
  geom_area() +
  geom_hline(yintercept = sustainable_limit, color = "red", linetype = "dashed") +
  labs(
    title = "Sectoral Contributions to Water Demand vs. Sustainability Threshold",
    x = "Year",
    y = "Total Gallons per Year",
    fill = "Sector"
  ) +
  scale_fill_manual(
    values = c("public_gal" = "skyblue", "tewu_gal" = "orange", "irrigation_gal" = "seagreen"),
    labels = c("Public", "Thermoelectric", "Irrigation")
  ) +
  theme_minimal()
```
This stacked area chart illustrates how three major sectors—public supply, thermoelectric power generation, and irrigation—contribute to total annual water demand in the United States from 2008 to 2020. Each colored layer represents a sector’s share, with public supply (green) as the largest contributor, followed by thermoelectric (blue), and irrigation (orange) as the smallest. The red dashed line marks the estimated sustainability threshold. In most years shown, total water demand either reaches or slightly exceeds this threshold, indicating that current consumption levels are already pressing against sustainable limits. The graph highlights the critical role of public and thermoelectric sectors in driving overall demand and underscores the need for targeted sector-specific water conservation and efficiency strategies to maintain long-term water security.




## Backtracking Population data
```{r}
pop_model <- lm(total_pop ~ poly(year_4, 2), data = plot_data)
backcast_years <- tibble(year_4 = 2000:2021)
backcast_data <- backcast_years %>%
  mutate(total_pop = predict(pop_model, newdata = backcast_years),
         sex = "Both", origin = "Total", race = "All")
plot_data_full <- bind_rows(backcast_data, plot_data)


ggplot(plot_data_full, aes(x = year_4, y = total_pop)) +
  geom_line(color = "darkgreen") +
  labs(
    title = "US Population Trend (2000–2100, Projected)",
    x = "Year",
    y = "Total Population"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```
Same population data from above accept the years 2000 - 2021 were added to find when exactly the sustainability threshhold was reached

## Adjusted Water Sustainability Graphs
```{r}
# STEP 1: Aggregate historical water use across all sectors
public_use <- water_use_long %>%
  group_by(year) %>%
  summarise(public_gal = sum(mgd, na.rm = TRUE) * 1e6 * 365)

tewu_use <- tewu_long %>%
  group_by(year) %>%
  summarise(tewu_gal = sum(tecuftot_mgd, na.rm = TRUE) * 1e6 * 365)

irrigation_use <- irrigation_long %>%
  group_by(year) %>%
  summarise(irrigation_gal = sum(mgd, na.rm = TRUE) * 1e6 * 365)

# STEP 2: Combine all three sectors into one table
water_total <- full_join(public_use, tewu_use, by = "year") %>%
  full_join(irrigation_use, by = "year") %>%
  arrange(year) %>%
  mutate(across(ends_with("_gal"), ~replace_na(., 0))) %>%
  mutate(total_gal_year = public_gal + tewu_gal + irrigation_gal)

# STEP 3: Fit linear model and project to 2100
water_model <- lm(total_gal_year ~ year, data = water_total)

future_years <- tibble(year = 2000:2100)

water_future <- future_years %>%
  mutate(total_gal_year = predict(water_model, newdata = future_years))

# STEP 4: Set sustainability threshold as max historical water use
sustainability_threshold <- max(water_total$total_gal_year, na.rm = TRUE)

# STEP 5: Identify first unsustainable year
unsustainable_year <- water_future %>%
  filter(total_gal_year > sustainability_threshold) %>%
  slice(1) %>%
  pull(year)

print(paste("⚠️ Water use exceeds sustainability threshold in:", unsustainable_year))

# STEP 6: Plot results
ggplot(water_future, aes(x = year, y = total_gal_year)) +
  geom_line(color = "orange", size = 1.2) +
  geom_hline(yintercept = sustainability_threshold, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Projected US Water Demand vs. Sustainability Threshold (2000–2100)",
    subtitle = paste("Water use becomes unsustainable in", unsustainable_year),
    x = "Year",
    y = "Total Water Use (Gallons per Year)"
  ) +
  theme_minimal()
```
This graph presents a long-term projection of total U.S. water demand from 2000 to 2100, based on aggregated data from public supply, thermoelectric, and irrigation sectors. Using a linear regression model trained on historical annual water use data, the yellow line forecasts future demand in gallons per year. The red dashed line represents a sustainability threshold, defined as the maximum total water usage observed in the historical record. The graph clearly shows that projected water use surpasses this threshold in the year 2018, marking the point at which demand exceeds what was historically sustainable. From that year forward, the gap between demand and the threshold continues to widen, suggesting that under current consumption patterns, the U.S. is on an unsustainable water trajectory. This visualization underscores the urgent need for water conservation, improved efficiency, and policy interventions to manage growing demand in the face of finite freshwater resources.








